\documentclass[11pt,a4paper]{article}
\usepackage[margin=2.5cm]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{fancyhdr}
\usepackage{enumitem}
\usepackage{parskip}

\setlength{\headheight}{14pt}
\definecolor{codebg}{RGB}{248,250,252}
\definecolor{codegray}{RGB}{100,100,100}
\definecolor{codeblue}{RGB}{37,99,235}

\lstdefinestyle{python}{
  backgroundcolor=\color{codebg},
  basicstyle=\ttfamily\footnotesize,
  keywordstyle=\color{codeblue},
  commentstyle=\color{codegray}\itshape,
  numbers=left, numberstyle=\tiny\color{codegray},
  numbersep=8pt, frame=single, rulecolor=\color{gray!30},
  breaklines=true, language=Python, showstringspaces=false
}

\hypersetup{colorlinks=true,linkcolor=black,citecolor=black,urlcolor=blue}
\pagestyle{fancy}
\fancyhf{}
\lhead{\small NeuFlow v3 --- Progress Report}
\rhead{\small April 2026}
\cfoot{\thepage}

\title{\textbf{Adapting an Implicit Neural Decoder for Optical Flow}\\[0.3em]
\large Two-week exploration: replacing NeuFlow v2's upsampler\\
with an InfiniDepth-style implicit decoder}
\author{Shriman Raghav Srinivasan}
\date{April 27, 2026}

\begin{document}
\maketitle\thispagestyle{fancy}

\section*{What this is}

Over the past two weeks I've been trying to swap out NeuFlow v2's convex upsampler with an implicit neural decoder, based on the approach in InfiniDepth (CVPR 2025). The idea seemed promising --- InfiniDepth showed that predicting depth at arbitrary continuous coordinates rather than on a fixed grid gives sharper boundaries and better fine-grained accuracy. I wanted to see if the same would apply to optical flow.

This isn't a complete project. It's a first attempt --- some things worked, some didn't, and the main bottleneck turned out to be training data rather than architecture. Here's what I did and what I found.

\section{Background: the two papers}

\subsection{NeuFlow v2}

NeuFlow v2~\cite{neuflowv2} is a real-time optical flow network (7.3M params) that avoids the expensive iterative refinement of RAFT by doing a single pass:
\begin{enumerate}[leftmargin=*,itemsep=1pt]
  \item CNN encoder extracts features at $1/8$ and $1/16$ resolution
  \item Cross-attention computes global correlation at $1/16$ (cheaper than 4D cost volume)
  \item GRU refinement runs 8 iterations at $1/8$
  \item Convex upsampler produces the final full-res flow
\end{enumerate}

The convex upsampler predicts the final flow as a weighted average of neighboring coarse flow values:
\[
  \mathbf{u}(x,y) = \sum_{i,j \in \mathcal{N}} w_{ij}(x,y)\,\hat{\mathbf{u}}_{ij}, \quad \sum_{i,j} w_{ij}=1,\; w_{ij}\ge 0
\]
The weights $w_{ij}$ are predicted by a small convolutional head. It's fast and works well but has an obvious limitation: the predicted flow at any pixel is always a convex combination of coarse values from a $3\times3$ grid. It can't predict anything sharper than what the $1/8$ coarse flow already captures.

\subsection{InfiniDepth}

InfiniDepth~\cite{yu2026infinidepth} represents depth as a continuous implicit function: for any $(x, y)$ coordinate, query the depth from feature maps rather than reading off a grid. This lets them run the decoder at 4K or 8K without retraining, and get sharper boundaries.

The core is their local implicit decoder. Given a query coordinate $(x,y)$, they bilinearly sample features from three ViT encoder layers to get local descriptors $f^1, f^2, f^3$ at different scales. These get fused hierarchically (Eq. 3 in their paper):
\[
  h^{k+1} = \mathrm{FFN}^k\!\left(f^{k+1} + g^k \odot \mathrm{Linear}(h^k)\right), \quad k = 1,\ldots,L-1
\]
where $g^k$ is a learnable channel-wise gate. Shallow features first, then progressively deeper ones. The final $h^L$ goes into an MLP that outputs depth.

They train for 800k steps on 8 GPUs with sparse L1 supervision --- randomly sample $N$ coordinate-depth pairs per image and supervise those, instead of the whole dense map.

\section{What I implemented}

The backbone of NeuFlow v2 is completely frozen. I only replaced the upsampling head. The new decoder:

\begin{enumerate}[leftmargin=*,itemsep=2pt]
  \item For any query pixel $(x,y)$, bilinearly samples features from three backbone maps:
    \begin{itemize}[itemsep=0pt]
      \item \texttt{ctx\_s8} --- 64-dim context features from img0 at 1/8 scale
      \item \texttt{feat\_s8} --- 128-dim cross-frame matching features at 1/8
      \item \texttt{feat\_s16} --- 128-dim features at 1/16
    \end{itemize}
  \item Fuses them using InfiniDepth's Eq.~3 twice (shallow $\to$ deep):
    \begin{align*}
      h^2 &= \mathrm{FFN}_1\!\left(\texttt{feat\_s8} + \sigma(g_1) \odot W_1\,\texttt{ctx\_s8}\right) \\
      h^3 &= \mathrm{FFN}_2\!\left(\texttt{feat\_s16} + \sigma(g_2) \odot W_2\,h^2\right)
    \end{align*}
  \item Also samples img1 features at the \emph{warped} location $(x + u_c,\, y + v_c)$ where $(u_c, v_c)$ is the coarse flow. This gives the decoder explicit cross-frame correspondence, which InfiniDepth doesn't need since depth is single-image.
  \item Concatenates everything and runs through an MLP: $[h^3 \mid \texttt{feat1\_warped} \mid (x,y)_\text{norm} \mid \mathbf{u}_c/WH] \in \mathbb{R}^{260} \to \Delta\mathbf{u} \in \mathbb{R}^2$
  \item Adds $\Delta\mathbf{u}$ to the coarse flow
\end{enumerate}

The output layer of the MLP is zero-initialized, so at the start of training $\Delta\mathbf{u} = 0$ and the model behaves exactly like v2. This prevents the gradient explosion that would otherwise happen when the decoder is completely random.

Total new parameters: 265K out of 7.3M.

\subsection{Training}

I trained with frozen backbone, lr $= 2\times10^{-4}$ for the decoder only, AdamW, batch size 4. Loss is sparse L1 on $N=4096$ randomly sampled query points per image, supervised by bilinearly interpolated GT flow --- same as InfiniDepth's Eq.~7.

The training set is VKITTI2 (2,121 pairs), using Scenes 18 and 20 as validation (1,174 pairs). This is the most significant limitation, which I'll get to shortly.

\begin{algorithm}[H]
\caption{Training loop (simplified)}
\begin{algorithmic}[1]
\State Load NeuFlow v2 weights; freeze backbone
\State Zero-init decoder output layer
\For{each step}
  \State Sample batch $(I_0, I_1, \mathbf{u}^*)$
  \State Run frozen backbone $\to$ coarse flow + feature maps
  \State Sample $N=4096$ query coordinates: 50\% flow-boundary-weighted, 50\% uniform
  \State Run decoder at query coords $\to$ $\hat{\mathbf{u}}(\mathbf{q})$
  \State $\mathcal{L} = \frac{1}{N}\|\hat{\mathbf{u}}(\mathbf{q}) - \mathbf{u}^*(\mathbf{q})\|_1$
  \State Backprop through decoder only; clip grad norm to 1.0
\EndFor
\end{algorithmic}
\end{algorithm}

\section{Results}

\begin{center}
\begin{tabular}{lcccc}
\toprule
& \textbf{Mean EPE} & \textbf{Median EPE} & \textbf{1px acc} & \textbf{3px acc} \\
\midrule
NeuFlow v2 (convex upsampler) & 2.23 px & 1.08 px & 46.2\% & 78.4\% \\
NeuFlow v3, step 10k & \textbf{3.15 px} & 1.87 px & 7.2\% & 71.1\% \\
\bottomrule
\end{tabular}
\end{center}

Not better than v2 yet. The decoder does learn something useful (EPE improves from random init), but it doesn't beat the baseline. The gap is almost entirely explained by dataset size.

\subsection{The overfitting problem}

Training past 10k steps makes things consistently worse:

\begin{center}
\begin{tabular}{cc}
\toprule Step & Mean EPE \\ \midrule
5k & 3.29 \\ \textbf{10k} & \textbf{3.15} \\ 15k & 3.47 \\ 20k & 3.16 \\ 25k & 3.56 \\ 30k & 3.69 \\
\bottomrule
\end{tabular}
\end{center}

At batch size 4 and 2,121 training pairs, step 10k is already $\approx$19 epochs. The model has memorized the training set. RAFT, NeuFlow, and basically every other optical flow method trains on FlyingChairs (22k pairs) and FlyingThings3D (21k pairs) before ever touching a real-world dataset. I skipped that entirely because of time and just trained directly on VKITTI2. That's the primary issue.

\subsection{End-to-end training didn't work}

I also tried unfreezing the backbone and training everything jointly with differential learning rates (backbone $10^{-5}$, decoder $2\times10^{-4}$). This completely failed --- the training EPE oscillated between 1px and 100px throughout, and the final model had basically zero accuracy (1px acc = 0\%). 

The reason is pretty obvious in hindsight: InfiniDepth runs 800k steps on 8 GPUs to get the backbone and decoder to co-converge. With 30k steps on one GPU and only 2k training pairs, the backbone features shift every step, the decoder chases a moving target, and nothing converges. Freezing the backbone gives stable features and the decoder can actually learn from them.

\section{What I'd do next}

The fix is straightforward: download FlyingChairs and FlyingThings3D and run the standard curriculum (Chairs $\to$ Things $\to$ VKITTI2). That should give the decoder enough data to actually learn and likely push EPE below the 2.23px baseline.

A few smaller things I'd also try:
\begin{itemize}[leftmargin=*,itemsep=2pt]
  \item Initialize the gate parameters to $-2$ instead of $0$ so $\sigma(-2)\approx0.12$ at the start. Currently gate init gives $\sigma(1)\approx0.73$, which means the context features contribute a lot from the first step with random projection weights. Starting more conservatively might help early training stability.
  \item Once the decoder converges on the full curriculum, try phase-2 fine-tuning: unfreeze the whole model at a very small lr ($5\times10^{-6}$) for another 20k steps. This is the standard way to do this without the divergence problem I hit.
\end{itemize}

\section*{Summary}

Two weeks in: I have a working implicit decoder attached to NeuFlow v2 that trains stably (when the backbone is frozen), understands the connection between InfiniDepth's approach and optical flow, and identified the main bottleneck (data). The architecture is fine. The next step is just getting the right training data, which is a matter of download time, not implementation.

Code: \url{https://github.com/shrirag10/Neuflowv3}

\begin{thebibliography}{9}
\bibitem{neuflowv2}
Zhao et al., \emph{NeuFlow v2: High-Efficiency Optical Flow Estimation on Edge Devices}, arXiv:2408.10161, 2024.

\bibitem{yu2026infinidepth}
Yu et al., \emph{InfiniDepth: Arbitrary-Resolution and Fine-Grained Depth Estimation with Neural Implicit Fields}, arXiv:2601.03252, CVPR 2025.
\end{thebibliography}

\end{document}
